Mendelian randomization with genome-wide association studies data

### 2022-04-27 reply-to-review1/mr-review.Rmd
updated bib file to match Rmd citations

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

ggplot2::ggsave("figures/mr-gwas-counts.png")

@@

@foley_clustering_2021 recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them. They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@foley_clustering_2021].

@@

# Mendelian Randomization with one Instrument

We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where 
there is a single genetic marker and a single outcome variable. Mendelian randomization with one 
instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
whether variation in the genotypes at the genetic marker causes variation in the 
"outcome" variable, which is typically a second complex trait. When the genetic marker 
is a SNP on an autosome, the cohort is partitioned into three 
classes, two homozygote classes and a single heterozygote class, based on genotypes 
for that marker. When the three assumptions of mendelian randomization are satisfied, 
estimation of the causal effect of the exposure on the outcome is possible.

The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
we focus on continuous outcomes before considering other classes of outcome variables in a later section.

## Assumptions of Mendelian Randomization

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].

The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

If the first assumption, about the association between the genetic marker and the exposure, holds, 
but this association is weak, then an amplification of biases, such as those due to violations of 
assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
the strength of association between the SNP and exposure. We later discuss, in the context of 
methods for multiple SNP instruments, how this finding, coupled with the recognition of 
the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].

The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
the need for the genetic marker to affect the outcome only through the exposure. In a 
causal diagram, this means that the only path from the marker to the outcome is that 
containing the exposure.

The third assumption ensures that the genetic marker is
independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
assumption, we can't verify this assumption with observational data, so analysts need to 
use tools like sensitivity analysis to ensure validity of findings.

To understand the origin of the term "Mendelian randomization", it is helpful to 
recall Mendel's inheritance laws [@mendel1865versuche]. Under the assumption that alleles segregate 
randomly from parent to offspring, the offspring genotypes are unlikely to be 
associated with confounders of the exposure-outcome relationship. Additionally,
reverse causation is unanticipated since germ-line genotypes are 
fixed at conception and, thus, precede realization of other observed variables. Reverse causation 
occurs when the outcome at an early time point has a causal effect on the exposure variable. While 
reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out 
several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].

## Statistical models and methods for MR with one instrument

Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
that leverage multiple SNP instruments.

### Ratio of coefficients method

The ratio of coefficients method, also known as the Wald method, estimates the 
causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. 
For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).

\begin{equation}
  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
  (\#eq:wald)
\end{equation}

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with 
summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its 
direct use in the GWAS setting.

### Two-stage methods

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

The resulting 
fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, 
where the outcome is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)). Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1). Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
  (\#eq:lm1)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
  (\#eq:lm2)
\end{equation}

Note that the uncertainty in the fitted values from 
the first regression is not considered when performing the second regression.
For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
incorrect in two-stage calculations. This and other observations led researchers to develop 
likelihood-based MR methods.

### Likelihood-based methods

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. We consider it below before introducing a Bayesian method.

Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
  (\#eq:liml1)
\end{equation}

\begin{equation}
  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
  (\#eq:liml2)
\end{equation}

Limited information maximum likelihood is sometimes called the maximum likelihood 
counterpart of two-stage least squares, and it yields the same causal estimate as 
two-stage least squares and the ratio method when used with a single SNP 
instrument. Additionally, the limited information maximum likelihood framework can accommodate $K > 1$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.

One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

\begin{eqnarray}
\begin{pmatrix}
X_i \\
Y_i
\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
\eta_i
\end{pmatrix}, \Sigma
\right)\nonumber \\
\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
\eta_i = \beta_0 + \beta_1 \xi_i
  (\#eq:b1)
\end{eqnarray}

For each subject $i$, the exposure and outcome values come from a bivariate normal 
distribution. The mean of the exposure distribution is assumed linear in 
the SNP instrument, and the mean of the outcome distribution is linear in the 
mean exposure [@burgess_review_2017;@jones_choice_2012]. 
@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
instruments, outperforms the limited information maximum likelihood model in terms 
of frequentist coverage levels.

### Semiparametric methods

Semiparametric instrumental variable methods, which feature parametric and nonparametric components, 
typically assume a parametric model connecting the exposure and outcome, but don't 
impose distributional assumptions on the errors [@burgess_review_2017]. Compared to 
fully parametric models, semiparametric models often are more robust to model 
misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow 
@burgess_review_2017 by discussing three semiparametric strategies, generalized 
method of moments, continuous updating estimator, and G-estimation of structural 
mean models.

Generalized method of moments can be viewed as a more flexible form of two-stage 
least squares that handles heteroscedastic errors and nonlinearity in the two 
regressions [@burgess_review_2017]. Before specifying the model, we introduce 
notation. $\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if 
we forced $X$ to take value $x$ for every subject [@pearl2009causality].

The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).

\begin{eqnarray}
  \sum_i (y_i - f(x_i;\beta)) = 0\nonumber \\
  \sum_i z_i(y_i - f(x_i;\beta)) = 0
  (\#eq:gmm)
\end{eqnarray}

The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$. @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).

Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].  
We follow @burgess_review_2017 by defining the potential outcome $Y(x)$ as the outcome value that we would 
have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
outcome is displayed in Equation \@ref(eq:smm).

\begin{equation}
\mathbb{E}(Y(x) - Y(0) | X = x, Z = z) = \beta_1 x
  (\#eq:smm)
\end{equation}

The causal effect parameter is $\beta_1$. @burgess_review_2017 derives the estimating equations for 
$\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is 
independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero 
covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).

\begin{equation}
\sum_i (z_{ik} - \bar z_{\cdot k})(y_i - \beta_1 x_i) = 0
  (\#eq:smm1)
\end{equation}

Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).

# Assumption violations and Sensitivity Analysis

It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations.
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}

library(dagitty)
library(ggplot2)

d1 <- dagitty('dag {
bb="0,0,1,1"
G [pos="0,0"]
W [pos="0.5,0"]
X [exposure, pos="1.0,0"]
Y [outcome, pos="1.5,0"]
G -> W
W -> X
W -> Y [pos="1,1"]
X -> Y
}')

plot(d1)

```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which affects Y. Unobserved confounder U affects both X and Y."}
d1 <- dagitty('dag {
bb="0,0,1,1"
G [pos="0.0,0.0"]
U [pos="0.75,0.5"]
X [exposure, pos="0.5,0.0"]
Y [outcome, pos="1.0,0.0"]
G -> X
U -> X
U -> Y
X -> Y
}'
)
plot(d1)

Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of 
possible assumption violations.

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ if the assumptions be violated.
Because not all confounding 
variables are known or measured, assumptions 2 and 3 are not fully verifiable 
in MR studies. Recognition of this fact has inspired the development of 
sensitivity analysis tools. While we can't assess whether a SNP instrument is 
associated with every confounder of the exposure-outcome association, it is 
possible to examine the associations of the SNP instrument with the measured 
covariates. While absence of such associations doesn't guarantee satisfaction 
of the assumption, presence of SNP-covariate associations must be investigated
carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].

@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when 
working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the
MR assumptions, to the extent possible, and performing robust analyses. For example, 
they used measured covariates to assess for possible associations with the SNP 
instruments. While they can't rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.

To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of 
C-reactive protein (CRP) levels on coronary artery disease risk with four genetic 
variants in the CRP gene region and 17 other genetic variants that affect coronary 
artery disease risk. They used measured covariates to probe for SNP-covariate 
associations, and followed it up with scatter plots and Cochran's Q test on the causal 
estimates to inquire about whether the 
SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the 
meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011]. Additionally, Egger regression can be useful in this 
setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].

# Mendelian Randomization with multiple SNP instruments

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in these studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to three considerations:

1. instrument selection (from among the available SNPs)  
1. modeling and controlling for horizontal pleiotropy  
1. statistical inference procedure

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping [@purcell2007plink].

However, LD clumping for choosing instrument SNPs can be problematic for several 
reasons. First, the chosen SNPs may not be causal for the outcome. This can 
happen, for instance, by using a noncausal SNP that is in high LD with a 
causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
causal SNPs may be in high LD together. LD clumping is likely to omit one or more
of the causal SNPs, which can diminish statistical power in MR analysis.

Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that 
bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
the standard MR assumptions and can lead to biased 
causal effect estimates and diminished statistical power. We discuss approaches to 
modeling and accounting for horizontal pleiotropy below.

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs.

Below, we discuss a number of MR methods that use multiple 
SNP instruments. The overall trend appears to be in the direction of increasingly sophisticated 
modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
attempting to maintain scalability and computational efficiency.

One approach to integrating multiple SNP instruments into a MR framework is
through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
calculate an unweighted score as the number of exposure-increasing alleles in 
the subject's genotypes. They also calculate a weighted allele score by using 
the exposure effects as weights. For example, a subject with $g_k$ copies of 
exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$. 
@burgess_use_2013 found that when weights are obtained from external data or from 
cross-validation or jackknife approaches applied to the analysis data, the allele score 
functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].

@kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to 
identify invalid SNP instruments [@tibshirani1996regression]. sisVIVE's 
advantage over earlier methods is that it doesn't require the analyst 
to know which SNP instruments are valid. Instead, it requires that at 
least 50\% of the instruments be valid. sisVIVE outperforms two-stage 
least squares in many ways and performs similar to oracle two-stage 
least squares. Simulations and data analysis results reveal that 
sisVIVE is robust to possibly invalid instruments.

@windmeijer_use_2019, building on the research from 
@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
after recognizing that sisVIVE misclassifies valid SNP instruments as 
invalid when the invalid SNP instruments have strong effects on the 
exposure. Consistent selection of invalid SNP instruments, they found, 
depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
proposed a median estimator with consistency that doesn't 
depend on SNP instrument-exposure association strength or the SNP instrument 
correlation structure. They then applied methods from @zou2006adaptive to 
achieve a consistent estimator with the same asymptotic distribution as the oracle 
two-stages least squares. One important limitation of this thread of research is that
both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.

@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their
work is the proposal to create a new instrumental variable as a linear combination of genotype data from 
a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically  
pleiotropic traits. For a collection of $K$ SNPs, they choose the $K$-vector that meets these criteria and 
maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a 
$\ell_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation 
of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. 
@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are 
unbiased in their simulation settings.

@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by 
environment interactions to detect and to adjust for bias due to horizontal pleiotropy. They leverage the insight 
that a SNP instrument-covariate interaction that induces differences in SNP instrument-exposure associations 
@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction. Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort. The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics:

1. Estimate the SNP-exposure and SNP-outcome associations at a range of covariate values  
1. Regress the SNP-outcome estimated associations on the SNP-exposure estimated associations  
1. Treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.

@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set 
of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant 
pleiotropy assumption.

MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the 
exclusion restriction assumption and building on existing G-estimation 
methods to create an estimator that is robust to violations 
of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting]. They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.

Finally, before considering methods for MR with GWAS summary statistics, we consider 
a new method for one-sample MR. @berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
one-sample MR with individual-level data, they construct their 
model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.

### Methods with Summary data

The proliferation of publicly available GWAS data coincides with the acceleration
of methods for two-sample MR with GWAS summary statistics. Our timeline highlights some
of the many two-sample MR methods with GWAS summary statistics, with an increasing 
density of methods over time in the last three years (Figure \@ref(fig:timeline)).

```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
knitr::include_graphics("figures/timeline.pdf")
```

@burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).

\begin{equation}
  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
  (\#eq:ivw)
\end{equation}

@bowden_mendelian_2015 adapted Egger regression, originally developed 
to detect bias in meta-analyses, to detect bias from horizontal 
pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data
[@egger1997bias]. Their approach is termed MR-Egger. They formulate their model as

\begin{eqnarray}
X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
(\#eq:egger)
\end{eqnarray}

where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.

@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).

\begin{eqnarray}
Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
= \Gamma_jG_{ij} + \epsilon'_{Yij}
(\#eq:egger1)
\end{eqnarray}

In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.

MR-Egger fits the regression model in Equation \@ref(eq:egger2).

\begin{equation}
\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
(\#eq:egger2)
\end{equation}

The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
the regression parameters are $\beta_{0E}$ and $\beta_E$.

The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].

>>>>

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: ../exported.bib
csl: ../elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

ggplot2::ggsave("../figures/mr-gwas-counts.png")

@@

@fang2021evaluating recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them. They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@fang2021evaluating].

@@

## Recent Findings
### Omnigenic MR
Recent research on the omnigenic hypothesis, which posits that every SNP’s effect on a trait is nonzero, has informed MR methods development [@boyle2017]. @wang_mendelian_nodate recognized the limitations of previous MR methods in the context of the omnigenic hypothesis. In response, @wang_mendelian_nodate developed a MR method that uses all genome-wide SNPs as instruments. Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive simulations, including those with model misspecifications, to characterize their method’s statistical power and robustness. Finally, they applied the new method to identify multiple complex traits that affect coronary artery disease and asthma. These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
### Sample overlap in biobanks
The issue of overlapping samples in the two-sample MR with GWAS summary data has recently received attention in the published literature. @fang2021evaluating, working with data from the UK Biobank, found that SNP instruments derived from overlapping samples explained a higher proportion of the variance compared to those derived from non-overlapping samples. They argue that block jackknife resampling MR enables causal inference while avoiding bias due to overlapping samples. @sadreev2021navigating analyzed 2,514 traits from the UK Biobank and evaluated the impact of winner’s curse on MR analysis. Winner’s curse, they found, amplifies weak instrument bias without inflating the false discovery rate. This finding led the authors to design a pseudoreplication process that reduces bias in MR studies. @berzuini_bayesian_2020 proposed a Bayesian approach for one-sample MR analysis where SNP instruments are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of one-sample MR with individual-level data, they construct their model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference proceeds by Markov chain Monte Carlo sampling. They also elaborate their model for a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.

### Bi-directional causal inference with MR to account for reverse causation
Reverse causation occurs when the outcome at an early time point has a causal effect on the exposure variable. While reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out several scenarios where this assumption may not hold [@davey_smith_mendelian_2003]. In recognition of this observation, @wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. GRAPPLE facilitates identification of SNP instruments that contribute to the additional profile likelihood modes. A researcher using GRAPPLE may then recognize a confounding factor for each mode. With additional GWAS summary data with traits that reflect the confounding factors, GRAPPLE can fit multivariable MR models when the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].

## Summary and Outlook

In summary, we have presented a comprehensive review on 47 methods for MR analysis in GWAS (Figure 4).  The overall trend of the methodology development for MR analysis is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs. We hope our detailed review would benefit both methodology developers and applied analysts to further advance the development of MR methods and aid in their applications towards large, biobank-scale datasets, thus offering the possibility of discovering even more causal relationships among complex traits [@bycroft_uk_2018].
